% This is sigproc-sp.tex -FILE FOR V2.6SP OF ACM_PROC_ARTICLE-SP.CLS
% OCTOBER 2002
%
% It is an example file showing how to use the 'acm_proc_article-sp.cls' V2.6SP
% LaTeX2e document class file for Conference Proceedings submissions.
% 
%----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V2.6SP) *DOES NOT* produce:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) Page numbering
%
%  However, both the CopyrightYear (default to 2002) and the ACM Copyright Data
% (default to X-XXXXX-XX-X/XX/XX) can still be over-ridden by whatever the author
% inserts into the source .tex file.
% e.g.
% \CopyrightYear{2003} will cause 2003 to appear in the copyright line.
% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
%
%
%---------------------------------------------------------------------------------------------------------------
% It is an example which *does* use the .bib file (from which the .bbl file
% is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission,
% you need to 'insert'  your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% Questions regarding SIGS should be sent to
% Adrienne Griscti ---> griscti@acm.org
%
% Questions/suggestions regarding the guidelines, .tex and .cls files, etc. to
% Gerald Murray ---> murray@acm.org 
%
% For tracking purposes - this is V2.6SP - OCTOBER 2002


\documentclass[12pt]{article}

\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{0in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\textwidth}{6in}
\setlength{\textheight}{9in}
\setlength{\parindent}{0in} 

\usepackage{graphicx} %For jpg figure inclusion
\usepackage{times} %For typeface
\usepackage{epsfig}
\usepackage{xcolor} %For Comments
%\usepackage[all]{xy}
\usepackage{float}
%\usepackage{subfigure} 
\usepackage{hyperref}
\usepackage{url}
\usepackage{parskip}
\usepackage{listings}

%% Elena's favorite green (thanks, Fernando!)
\definecolor{ForestGreen}{RGB}{34,139,34}
\definecolor{BlueViolet}{RGB}{138,43,226}
\definecolor{Coquelicot}{RGB}{255, 56, 0}
\definecolor{Teal}{RGB}{2,132,130}
%Uncomment this if you want to show work-in-progress comments
\newcommand{\comment}[1]{{\bf \tt  {#1}}}
% Uncomment this if you don't want to show comments
%\newcommand{\comment}[1]{}
% for color names see https://www.overleaf.com/learn/latex/Using_colors_in_LaTeX
\newcommand{\emcomment}[1]{\textcolor{ForestGreen}{\comment{Elena: {#1}}}}
\newcommand{\todo}[1]{\textcolor{blue}{\comment{To Do: {#1}}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\pagestyle{plain}
%
% --- Author Metadata here ---
%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\setpagenumber{50}
%\CopyrightYear{2002} % Allows default copyright year (2002) to be
%over-ridden - IF NEED BE. 
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
%(X-XXXXX-XX-X/XX/XX) to be over-ridden. 
% --- End of Author Metadata ---

\title{15 years of Sorting Competitions.}
%\subtitle{[Extended Abstract \comment{DO WE NEED THIS?}]
%\titlenote{}}
%
% You need the command \numberofauthors to handle the "boxing"
% and alignment of the authors under the title, and to add
% a section for authors number 4 through n.
%
% Up to the first three authors are aligned under the title;
% use the \alignauthor commands below to handle those names
% and affiliations. Add names, affiliations, addresses for
% additional authors as the argument to \additionalauthors;
% these will be set for you without further effort on your
% part as the last section in the body of your article BEFORE
% References or any Appendices.

\author{
Elena Machkasova\\
Computer Science Discipline \\
University of Minnesota Morris\\
Morris, MN 56267\\
elenam@morris.umn.edu
}
\date{}
\maketitle
\thispagestyle{empty}

\section*{\centering Abstract}
15 years ago, at MICS 2011, I presented a paper on a newly developed multi-part assignment for UMN Morris CSci 3501 “Algorithms and Computability” class: a sorting competition. This assignment is still a part of the course, with a novel challenge every year. The assignment was created to give students experience with optimizing an algorithm not just for theoretical Big-O efficiency, but for the speed of a concrete implementation in a programming language. The competition measures the time of the sorting part of the program (excluding input and output); the time of the programs that didn’t sort in the correct order was not counted. The programming language is Java since it’s the language used in a prerequisite of the course, and therefore all students were familiar with it. The competition went very well in terms of the students’ interest, creativity, and the ability to explore both the algorithms and the Java platform. 

The competition became a tradition, and most years was open to submissions from alumni, in addition to students in the class. 
The multi-part course assignment includes a presentation, and students in the class have an option of presenting on solutions from alums, which enriches the learning experience since such submissions are often based on real-world programming expertise. 

In this paper I discuss approaches to choosing a good competition problem, the setup and the rules, and the steps of the assignment. Most importantly, I hope the paper conveys the excitement of learning algorithms not just through theoretical exercises, but through real, unpredictable work with interesting non-textbook data and the nuances of a real world programming platform – the excitement that makes students come back for more sorting fun even after graduation. 


\newpage
\setcounter{page}{1}
\section{Introduction}\label{sec:intro}
In the process of earning a bachelor's degree in computer science students typically learn a variety of sorting algorithms, from bubble sort to radix sort and heap sort. They implement several of these algorithms and learn their $O$ running times, advantages, and disadvantages. However, very few courses require students to implement sorting of any data types other than integers, and perhaps strings. Thus, students often graduate without actual experience of adapting algorithms to more complex data types and sorting criteria. They also don't get a clear sense for what $O$ running times really mean (for example, that $O(n \log n)$ may be more efficient than in practice than $O(n)$, depending on the constants.) 
Getting a hands-on experience with developing a fast sorting algorithm would allow students to understand the tradeoffs and the contexts of applying different algorithms, as well as the art of implementing an algorithm in a programming language, such as choosing more efficient data types and data structures.    

Recognizing the need for such an assignment, I added the \textit{Sorting Competition} in the 2010 offering of UMN Morris CSci 3501 ``Algorithms and Computability'' course~\cite{mics2011}.  This is a 5-credit course, covering the first two chapters (and a few selected topics) of~\cite{CLRS}, followed by the theory of computation material. The sorting competition starts after Chapter II of~\cite{CLRS} is covered and includes multiple parts:
\begin{enumerate}
\item At least two rounds of preliminary competitions
\item The final round of the competition
\item Correctness analysis
\item Presentations
\item Final reports
\end{enumerate}
The entire assignment takes about a month and usually takes place in labs while more theoretical computability material is covered in lectures. 

The competition is set up as follows: students are given a sorting task (for example, sorting positive integers by the sum of their unique prime factors). Their goal is to design and implement the fastest algorithm in Java. There are some specific restrictions which I detail in Section~\ref{subsec:java-setup}.
\emcomment{And possibly a section on the general setup, such as the timing?}
The timing uses Java {\tt time} function and is applied only to the sorting itself, not reading in the data or printing out the output. Students are given a \textit{reference implementation}: an implementation of the sorting that provides a correct order, not optimized for speed. The output of the reference implementation is compared to students' sorted data to check if the sorting is correct. This allows students to experiment with different inputs and check their correctness.  

Over the years the competition became a tradition that extends to more than just students in CSci 3501: the challenge is open to anyone interested in participating, including faculty, alumni, and students not in the class. It often features one or two ``external'' submissions, i.e. submissions from those not in the class. Some alumni participate fairly regularly. The presentations are also open to everyone. This adds a social aspects to the competition, further promoting interest in algorithmic problem solving. I am really fortunate that my colleague Dr. Peter Dolan who teaches the CSci 3501 class in some years is just as excited about the competition as I am, and we often collaborate on putting together the challenge, regardless of who of us happens to teach the class in a given year.  

In the remainder of the paper I will detail the rules and the setup in Section~\ref{sec:rules}, list some of the challenges and the takeaways in Section~\ref{sec:past}, and discuss the role of the sorting competition in Morris CSci community in Section~\ref{sec:community}. Conclusions and future directions are discussed in Section~\ref{sec:conclusions}.

\section{Rules and setup}\label{sec:rules}

\subsection{General setup and starting files}\label{subsec:general}
The goal of the competition is to come up with the fastest sorting algorithms. The speed is measured from the start of the sorting to its finish, not counting reading in the data and writing it out. This is done by using Java \texttt{System.currentTimeMillis()} method to record the times right before the sorting starts and right after it ends. The input is read from a file before the timing starts, passes the unsorted array to a timed sorting method, and the sorted array is written out to a file after the sorting is done. 
Unix \texttt{diff} command may then be used to compare the output of the students' sorting to the correct order of the data. The input/output code is provided, and there are only limited changes that can be made to it (see Section~\ref{subsec:allowed-features} for details). 

In the first couple of years of the competition students had to write their own code to check if the data is sorted correctly. However, I found out that it is more beneficial to provide a \textit{reference implementation}: an implementation of sorting that provides the correct result, but is not optimized for speed. In fact, I often provide an implementation that deliberately includes inefficiencies, such as unnecessarily repeated computations, to give students a ``low-hanging fruit'' for optimization. The reference implementation often calls a Java predefined sorting method, such as \texttt{Arrays.sort()} (TimSort since \emcomment{fill in the year}, with a comparator that implements the comparison criterion for the given data. Since many students aren't familiar with this method, it provides an additional pedagogical benefit of demonstrating the use of predefined sorting methods with a customized comparator. 

In the recent years I also started providing a data generator. There usually are data parameters that can come from specified ranges, and data often follows specific distributions. Statistical properties of data can be important, and many top-scoring algorithms adjust to specific parameters (for example, they may be different for shorter strings than for longer ones). Providing a data generator allows students to study the problem space on their own. 

An important feature of the competition is that the sorting program is ran pinned to a single core using Unix \texttt{taskset -c 0} command, so parallelization is not an optimization option. The reason for this restriction is to keep the competition within the scope of the material covered in the first two chapters of~\cite{CLRS} and to avoid giving unfair advantage to a few students who are familiar with this technique. 

\subsection{Allowed code modifications}\label{subsec:allowed-features}
The reference implementation also serves as a template for students and clearly indicates what code can and cannot be modified. Most of the students' own code must be in the timed sorting method. However, there are a few other options for modifications. There are also important constraints. Below is the list of allowed modifications and their constraints.
\begin{enumerate}
\item Depending on the problem, students may or may not be allowed to add some information gathering as the program is reading the data. However, if it is allowed,  there are limitations:  the total memory to store this information is constant w.r.t. the size of the data. Usually there is a maximum amount of memory that can be used for this purpose. This means that computing some general information, such as the minimum and the maximum value, would be fine, but recording some information about each element would not be.
\item The sorting method must take the array of elements as produced by the reading portion of the template. Typically the sorting method in the reference implementation sorts in place, and therefore is void. However, students are allowed to change its return type to return a sorted array, and its type doesn't have to be the same as the given array. In this case students must provide their own method for writing out the resulting array to a file. Recall that the method is called outside of the timed portion of the program. The constraint is that no processing is done as the data is being written out. For example, if the result needs to be in binary and of a fixed length, it's not permitted to store the result as a decimal and convert to binary with padding as the results are written out. 
\item  There are also restrictions on the use of global variables. For example, in recent years a typical restriction was to allow global constant, but not variables. There are also limits on how large the total memory for constants can be. 
\item Since the program runs on a single core, the use of \textit{multithreading} is not prohibited because it doesn't provide the benefit of parallelization.  Interestingly, in 2022 one of the teams used multithreading and discovered that it improved their time even on a single core (see Section~\ref{sec:past} for details). Thus the rules were updated to explicitly allow multithreading, but limit the number of threads to four and make it constant, not depending on the data.
\end{enumerate}

\subsection{Java setup}\label{subsec:java-setup}
The sorting competition runs in the CS lab that's a Unix lab with the OpenJDK Java, currently OpenJDK 64-Bit Server VM, version 17. In order to avoid delays due to NFS access, the files are copied into a \texttt{/tmp} directory of one of the lab machines. 

Java Virtual Machine (JVM) provides a large number of automatic optimizations.  However, these optimizations are implemented in the Just-in-Time (JIT) compiler and require a method to be executed a large number of times before it is optimized. This is known as \textit{JVM warmup}~\cite{warmup}. In order to take full advantage of the JVM warmup, the template runs the sorting twice: once before timing, to allow the JVM to fully optimize the methods, and then with the timing, to measure the optimized sorting. The input to the second sorting is the cloned unsorted array. 

Moreover, in order to make sure that Java garbage collection from the first run doesn't interfere with the second, timed, run we pause the main thread for 10 milliseconds via \texttt{Thread.sleep(10)} before starting the timed run. This approach reduces unpredictability of the JVM optimization, memory allocation, and garbage collection, and it stabilized the times significantly, especially for sorting with larger memory consumption. 

%\emcomment{Mention clearing out all allocated memory}
Since the sorting is performed twice within each run of the program, it is essential that all memory allocated in the sorting is completely freed at the end, to make sure that nothing gets reused from the previous sorting. This requirement is an essential part of the correctness check (see Section~\ref{subsec:correct}). 

\subsection{Scoring and competition rounds}\label{subsec:rounds}
There may be occasional system tasks running on the machine during the competition. Therefore each program is ran three times on each dataset (restarting the JVM every time), and the median value is taken to be the result. It's not unusual to encounter a time outlier, especially with large memory allocation, but the median approach provides stability. For example, out of these three results of running the same program (in milliseconds), 9193, 10039, 9551, the one that counts for the competition is 9551. 

Every competition has at least three rounds (at least two preliminary and one final), each round featuring at least two different datasets. The datasets in the same round often differ in the number of elements to sort and in some other parameters of the data generator (see Section~\ref{sec:past} for examples). 

Students typically work in teams of two. Every team gets a team number; their code is written in a file \texttt{GroupN.java}, where \texttt{N} is replaced by their group number. 
\texttt{Group0.java} is the reference implementation, and it's ran first. 
Up until the final competition the groups maintain their anonymity, known only by their group number, and only their compiled code (the \texttt{.class} files), but not the source Java code, is made public. Making the \texttt{.class} files public allows students to test their programs against solutions by others without revealing anyone's source code. 

Group 0 (the reference implementation) runs first and outputs the sorted dataset into a file. The file is then used by a Unix \texttt{diff} command to compare the sorted file produced by every group to the output of Group 0. If they match, the sorting is correct, otherwise it's not. Space characters are ignored in the \texttt{diff}. 

The score in a round is determined as follows:
\begin{itemize}
\item The sum of places on all of the datasets in the round. For example, if a team was second on the first dataset and third on the second one in a round with two datasets, the sum of places is 5. 
\item If the sum of places in the same for two teams, the sum of medians of all datasets decides the tie. 
\item If a group sorts incorrectly, a penalty of 1000000 milliseconds is added for each incorrectly sorted dataset, pushing incorrect solutions towards the end of the list.
\end{itemize}
Below is the outcome of one of the preliminary rounds in 2023. Groups 1 and 4 sorted both datasets incorrectly, so they appear at the end of the list. Group 0 (the reference implementation) is still outperforming most of the submissions. That is because the focus at preliminary rounds is often on correctness, not yet at the optimizations. It's also worth noting that Group 3 sorted one dataset correctly, and the other one incorrectly. Since its time in the correctly sorted run was quite good (they actually were 1st, as the log indicates) it still ended up in 4th place.    
\begin{verbatim}
group 2 took place  1. 
The sum of places is 3, the sum of medians is 259.0
group 0 took place  2. 
The sum of places is 5, the sum of medians is 6900.0
group 6 took place  3. 
The sum of places is 7, the sum of medians is 7308.0
group 3 took place  4. 
The sum of places is 7, the sum of medians is 1000021.0
group 5 took place  5. 
The sum of places is 9, the sum of medians is 36843.0
group 1 took place  6. 
The sum of places is 11, the sum of medians is 2000000.0
group 4 took place  7. 
The sum of places is 14, the sum of medians is 2000000.0
\end{verbatim}

The final round is similar to preliminary rounds in that it tests the programs on two or three different datasets and uses the same scoring. 
However, as the submissions are pushed to github right before the final competition, the source code is posted and the members of each group are revealed. 
The place earned at the final round is the official place in the competition. 
A week or two after the competition (and after the correctness analysis, see Section~\ref{subsec:correct}), students present the summary of their algorithm. 
Their presentation slides are also posted on github. 

\subsection{Correctness check}\label{subsec:correct}
In addition to producing correct output on the testing data, submissions need to follow all the rules (see Section~\ref{subsec:allowed-features}) and correctly sort any data that can possibly be generated by the data generator. In order to provide a careful correctness review, once the competition is finished and all source code is posted, each group is assigned another group to scrutinize their code for correctness. 
A group that violated the rules in any way loses their place in the competition, even if their sorting was correct on the final round data. 

Additionally there is extra credit for reporting correctness issues in \textit{anyone's submission} before their reviewing group provides their review (or if it fails to notice the issue). 
In an interesting turn of events, there were at least two cases when a group discovered a correctness issue in their own code and chose to report it before their reviewing group noticed it. 
As per the rules, they lost their place in the competition, but earned a bit of extra credit for reporting an issue. 
External submissions (see Section~\ref{sec:community}) are also reviewed for correctness due to the extra credit opportunity for a team or a student who find out the issue. 

\section{Past challenges and takeaways}\label{sec:past}
The 15 years of the sorting competition featured a variety of problems, some offered by me and some by my colleague Dr. Dolan who taught approximately 1/3 of all course offerings. He didn't utilize public github repositories for his challenges the way I started doing in 2014, so I don't have access to them. However, even a subset of these challenges that's posted on my github provides a good overview of the problems and solutions. 

Dr. Dolan and I often collaborated on developing the challenges, regardless of who of us taught the class. The way to determine if a challenge was ``good'' was to start tossing ideas around about potential approaches to sorting - and if we identified multiple ones as something to explore, rather than definitely agreeing on the best one, the challenge was considered to be a good one. 
We also carefully cross-checked the problem statement to make sure that the ordering was total, with no ambiguity.  

\subsection{Selected past challenges}\label{subsec:list}
%\emcomment{list some, not all; not all on github; omitting some details}
Below I list the sorting competition challenges that posted in my github repositories. Since these are all past challenges, all of the final code and the results are posted. The code is located in the \texttt{src} directory that also contains the data generator, if available, and the reference implementation Group 0. 
The \texttt{README.md} file describes the problem, provides links to the results of all runs (preliminary and final), the results, and has links to slides from students' presentations slides. 

In the list of challenges below the specifics of data generation and parameters, as well as some insignificant details of the sorting criteria, are omitted. The sorting is in increasing order, unless specified otherwise. The sorting tasks were as follows:
%\emcomment{How to find things in a repo}
\begin{itemize}
\item 2014: strings of different length, to be sorted first by length, then by the number of 1s in their binary representation, and if any are strings are equal by the first two criteria, then lexicographically~\cite{sorting2014}.
\item 2016: 2D points on a grid with a ``timestamp'', generated as ``walks'', i.e. the next points separated by a small ``step'' from the previous, to be sorted by the distance to the closest of two separately generated constant points, with the timestamp serving as a tie breaker~\cite{sorting2016}.
\item 2017: Positive integers to be sorted by $P(N)$ defined as the product of two smallest different prime factors or the only prime factor of $N$ if it only has one. If $P(N_1) = P(N_2)$, $N_1$ and  $N_2$ are compared as regular numbers~\cite{sorting2017}.
\item 2021: non-uniformly distributed numbers (with the probability decreasing for larger numbers) to be sorted by the number of 1s in their binary representations, then by the length of the longest substring of 0s and 1s that appears at least twice in their binary representation, with the actual value of the numbers as a tie-breaker~\cite{sorting2021}. 
\item 2022: numbers that are represented as decimal fractions and rational numbers are to be sorted  by value.  For the numbers of equal value the decimal representation is considered ``smaller'' than any fraction for positive numbers and ``larger'' for negatives. The resulting sorted results must be written to a file preserving the original format~\cite{sorting2022}. 
\item 2023: Fixed length integers (padded with zeros) are to be sorted in increasing order by the sum of all unique prime factors. Numbers with the sums being equal are ordered by value in decreasing order~\cite{sorting2023}.
\item 2025: Binary strings of the same length are to be sorted by their distance from a randomly generated ``target'' string. Strings of equal distance are compared by the value of the number they represent~\cite{sorting2025}.
\end{itemize}
Some of the challenges have some common themes. For example, the 2017 and 2023 challenges both deal with primary factor of numbers, and the 2016 and 2025 are both based on distances from a specified constant or constants and have some similarity in data generation. 
However, similarity to past years' problems is not a concern since they are separated by several years, so there cannot possibly be an overlap of students. We also observed that small details, such as whether the length of the input strings is fixed, the data distribution, etc. make a difference in what algorithms produce the fastest solution, so even seemingly similar problems present unique challenges. 

The competition problems seem convoluted, but this leads to more interesting solutions and a wider variety of approaches. I don't have room to go over solutions for all challenges,
so I will pick two, 2022 and 2023 (see the descriptions of the sorting problem above), and discuss the algorithms and data representation that were the most efficient for these cases. 
 
\subsection{2022 challenge}\label{subsec:2022}
As mentioned above, the sorting involves decimal and rational fractions that are sorted by value, but are required to preserve the original format, such as $0.5$ or $\frac{3}{6}$ and have rules for resolving orderings of the numbers that have the same value, but different formats (see~\ref{subsec:list} and~\cite{sorting2022} for details).  

The top solution for this challenge utilized the approach that split the rationals and the decimals into two separate arrays and then further into ``buckets'', sorting each bucket using its own sorting algorithm: Java-provided TimSort for rationals and radix sort with counting sort for decimals. At the end all of the buckets and the two types of numbers are merged together, just like they would be in merge sort. The solution utilizes a sophisticated data representation that stores the numbers in the data size they fit in: Java \texttt{double} if possible and the special class BigRational for numbers that require more precision than double provides. The comparison between different types of numbers employs a variety of combinations of cases. 

The second place solution approached the problem quite differently: they split the data into positive and negative numbers. Within each subset they implemented a variant of bucket sort with a large number of buckets (based on the~\cite{CLRS} bucket sort), each bucket sorted using insertion sort. Interestingly, the positive and negative subsets were sorted by two different Java threads, and even though the program was pinned to a single core (see~\ref{subsec:allowed-features}), multithreading provided some improvement, presumably due to memory caching and CPU-level optimizations. This solution didn't need to merge the resulting buckets, just to concatenate them.  However, insertion sort, even on small buckets, was a bit slower than the winning group's solution. 

One group didn't go as far as the winning two in terms of developing their sorting algorithms, but put a lot of effort into optimizing the comparator, while keeping the original Java-provided TimSort. In particular, one group took advantage of the difference in sign and immediately provided the comparison results for a positive and a negative number. Another group presorted the given array by string length in a hope of creating a partially sorted array before passing it to TimSort since TimSort takes advantage of partial sorting. A yet another group implemented its own customized quicksort. 
There were also two groups that tried to optimize the comparison, but ended up not sorting correctly on the final competition. 

Overall, the challenge allowed for a variety of solutions that included different sorting algorithms and their combinations. The main challenge was to avoid unnecessary overhead during the comparison, and yet to be able to preserve the precision for all cases and the original format of the data. 

\subsection{2023 challenge}\label{subsec:2023}
In this challenge fixed length integers (padded with zeros) are to be sorted in increasing order by the sum of all unique prime factors. Numbers with the sums being equal are ordered by value in decreasing order. See~\cite{sorting2023} for details. 

This competition, just like the one in 2022 and many others, favors approaches that avoid unnecessary  computations. Note that there are two aspects of optimization: finding the prime factors and the actual sorting. Different groups focused on different aspects of the challenge. 

The winning group used a hashmap to store the sums of prime factors that they already computed, keyed by the number for which they computed it. This allowed them to reuse already computed prime factors if they encountered these numbers again in the process of dividing a new number by its prime factor. The approach is similar to memoization. The comparison method looked up the sum of prime factors of the two numbers (and if not stored yet, computed them and added to the hashmap), and if the two were equal, would resolve the difference by comparing the values.

The second place group had a reasonably efficient way of computing prime factors without storing them, then split the data into buckets by the sum of prime factors (there were a lot of duplicates), and then sorted each ``bucket'' using Java \texttt{Collections.sort()} method. 

One group used the Sieve of Eratosthenes for computing prime factors. However, some of the given numbers are very large, and this approach doesn't result in the fastest algorithm. There were other groups that explored the idea of storing the already computed sums of prime factors in a hashmap, but ran into correctness issues.

This computation also featured a submission from a student not in the class (a senior) whose algorithm was correct and the fastest (however, external submissions aren't counted as a benchmark for expectations for the class). This solution took advantage of storing the entire data item in Java \texttt{long}, where the first 32 bits store the sum of the prime factors and the second 32 bits store 
$-N$ since the sorting by value is done in reverse. This allowed the student to use predefined stream sorting methods. 

One thing to note is that the ability to store data in Java primitive type gives a very significant improvement. Not all challenges and not all cases allow for this optimization, but if it's possible, it is almost certain to be among the fastest solutions. 

\section{Competition as community building}\label{sec:community}
%\emcomment{Presentations, alumni participation}
Over the years the sorting competition became a tradition that goes beyond the CSci 3501 class. The final round, as well as the presentations, are open to anyone, and often gather a group of CSci faculty and students from other classes. 

Moreover, the competition itself is open to anyone interested in participating. 
The announcement about the competition is typically sent to the mailing list that includes current students and alumni. 
Participants who are not in the class have an option of revealing their name or remain anonymous, identified only by the group number (their code is posted with everyone else's, however)
Several alumni, some from more than 10 years ago, participated more than once. 
It's not unusual for these submissions to do very well -- some of these former students have worked in Java for years, including specifically on program optimization, and some are putting in an effort just for fun.
This past fall, in 2025, we had three submissions from alumni - a very exciting development!

Submissions from outside of the class aren't limited to alumni, though. In the past we've had seniors submit their solutions. We also once had a sophomore who decided to give it a try before taking the class, and did fairly well (although didn't win). 

Students in the class are given an option to do a presentation on an out-of-class submissions (since the authors aren't available to present), especially the ones that did quite well, for extra credit. This is an opportunity to explore algorithms and program optimization done by experts, and is a very valuable experience. 

It is great to see the excitement about a class assignment that keeps connecting alumni with students, even if indirectly. Establishing a tradition that survives and thrives for 15 years is not a small thing. 

\section{Conclusions and the future}\label{sec:conclusions}
Considering 15 years of running the sorting competition, there are a few technical takeaways:
\begin{enumerate}
\item Data representation matters. In a language like Java that has a significant overhead for using objects the ability to store data in primitive types (ints, doubles) is a great benefit. Other data representation advantages include storing precomputed fields in a custom object, such as the sum of prime factors in 2023 competition.  
\item Fixed length data is a good opportunity for radix sort. Finding the right tradeoffs between the number of paths of radix sort and the range each of the paths covers is important. 
\item Quicksort is a fast, in-place algorithm that can be modified to deal efficiently with a large number of duplicates if the problem includes those. However, it's not stable, so using it in a combined sorting should be done carefully.  It is common for quicksort to appear within top ten solutions (although the two detailed in Sections~ref{subsec:2022} and~\ref{subsec:2023} are somewhat of an exception due to the specifics of the problems). The other option of dealing with duplicates is to split data into ``buckets'' and sort each ``bucket'' separately, then combine.
\item Taking advantage of data structures and methods already defined in Java can be very beneficial.
For example, hashing results of a computation helps to avoid repeated long computations, e.g. prime factors. TimSort is a good sorting algorithm that allows students to focus just on optimizing a comparison.  
\end{enumerate}
In addition to purely technical takeaways there are other lessons brought on by the competition: it promotes team work and communication, it teaches students to communicate technical ideas and to use resources (there is nothing wrong with studying existing algorithms and modifying an existing implementation - there are plenty of details that need to be worked out in the process). 

Keeping the tradition of the sorting competition also works as community-building tool, fostering the culture of creative problem solving. The instructors need to be careful with AI-proofing the challenges (there should be no easy way to immediately implement the solution, although using AI resources as a supplement is certainly an option). Given the intricacy and the novelty of the data and the sorting criteria, I believe that even with AI resources the problem still fosters analytical skills and attention to implementation details. 

I hope the competition continues for as long as the course exists. 

\section{Acknowledgments}\label{sec:ack}
I am grateful to Dr. Peter Dolan for sharing the excitement about the competition and keeping it going in years when he teaches the class, as well as for numerous enlightening and fun discussions of potential competition challenges. I am also thankful to alumni who provided ideas for some of these challenges. Finally, my great appreciation goes to all participants, students and alumni, who bring in problem solving skills, creativity, and hard work to explore this exciting area of computer science.  


\bibliographystyle{acm}
\bibliography{MICS2026Sorting}

\end{document}

